---
title: "Getting started with SubtypeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with SubtypeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`SubtypeR` assigns a molecular subtype to every cell in a single-cell RNA-seq
dataset using UCell gene-signature scoring — the same way `cytotrace2()`
assigns a stemness score to every cell. Starting from a **raw count matrix**,
the pipeline:

1. Builds/uses a Seurat object and runs `SCTransform` normalization, keeping
   **all genes** (`return.only.var.genes = FALSE`) so signature genes are
   never dropped for not being highly variable.
2. Scores a panel of gene signatures per cell with `UCell`.
3. Assigns each cell a single subtype label using a per-signature quantile
   threshold (default: 75th percentile / Q3):
   - passes exactly one signature -> that signature's name
   - passes more than one -> `"Mixed"`
   - passes none -> `"Unknown"`

Two modes are available.

## Mode 1: Automatic (built-in HGSOC panel)

Use this when you're working with high-grade serous ovarian cancer (HGSOC)
and want to use the package's built-in, curated signature panel
(Differentiated / Immunoreactive / Mesenchymal / Proliferative).

```{r}
library(SubtypeR)

# raw_counts: genes x cells raw count matrix (dgCMatrix or matrix)
res <- RunSubtypeR(
  input       = raw_counts,
  mode        = "auto",
  cancer_type = "HGSOC"
)

table(res$results$Subtype)
head(res$results)

# QC plots
plot_subtype_scores(res$results)
plot_subtype_composition(res$results)

# Subtype is also written into the Seurat object's metadata
res$seurat_object$Subtype[1:5]
```

## Mode 2: Manual (any cancer type via your own .gmt file)

Use this for any other cancer type, or any custom set of gene signatures.
Supply a standard `.gmt` file: one signature per line, tab-separated, as
`name<TAB>description<TAB>gene1<TAB>gene2...`.

```{r}
res <- RunSubtypeR(
  input    = raw_counts,
  mode     = "manual",
  gmt_file = "my_signatures.gmt"
)

table(res$results$Subtype)
```

You can also parse a `.gmt` file yourself and inspect it before running:

```{r}
sigs <- read_gmt("my_signatures.gmt")
names(sigs)
sigs[["MySignature1"]]
```

## Starting from an existing Seurat object

If you already have a Seurat object with raw counts in its default assay,
pass it directly as `input` — `SubtypeR` will skip `CreateSeuratObject()` and
run `SCTransform` on it directly:

```{r}
res <- RunSubtypeR(seurat_obj, mode = "auto", cancer_type = "HGSOC")
```

## Tuning the subtype call

The Q3 (75th percentile) threshold can be changed via `assign_quantile`. A
higher value is more stringent (fewer cells called into a defined subtype,
more `"Unknown"`); a lower value is more permissive:

```{r}
res_strict <- RunSubtypeR(raw_counts, mode = "auto", assign_quantile = 0.9)
res_loose  <- RunSubtypeR(raw_counts, mode = "auto", assign_quantile = 0.5)
```

You can also re-run just the assignment step on existing scores without
re-scoring:

```{r}
new_calls <- assign_subtypes(res$results[, grep("_UCell$", colnames(res$results))],
                              quantile_prob = 0.9)
```
